Gelation as a condensation frustrated by hydrodynamics and mechanical tension

### 2012-07-05 gel_cond/trunk/paper.tex
maximum bond length

We use a novel experimental set-up that allows the observation of the entire process of gelation at a single particle level (see Methods). The principle is to introduce salt through a membrane in a charged colloids and non adsorbing polymers suspension (see Fig.~\ref{fig:methods}a). The polymers induce a very short range depletion attraction between colloids, however long range electrostatic repulsion between the colloids prevent any aggregation. When we introduce salt the charges get screened - repulsion is `turned off' - and attraction suddenly become dominant, eventually leading to the formation of a gel. The whole process takes place on the microscope stage and can thus be observed in three dimensions from the very beginning.

Figure~\ref{fig:methods}c-h shows a computer reconstruction from experimental coordinates of a typical gelation experiment. We are able to follow the state of the system from a Wigner crystal (Fig.~\ref{fig:methods}c) where the particles are far apart due to repulsion to a fully percolated network (Fig.~\ref{fig:methods}h).

Our protocol is not equivalent to a redispersion of an existing gel at the same state point either by shaking or shear melting~\cite{lu2008gelation,Teece2011,Bartlett2012}. It is closer to a quench and to application relevant processes where the sudden addition of a component (e.g. salt, acid, depletion agent) induces gelation in an otherwise stable suspension (e.g. milk).

@@

The data were collected on a Leica SP5 confocal microscope, using \unit{532}{\nano\meter} laser excitation. The temperature was controlled on both stage and objective lens, allowing a more precise density matching. The scanned volume is at least \unit{82 \times 82 \times 85}{\micro\metre}. Standard particle tracking techniques allow us to localise our particles with errors below $1\%$ of a diameter. We carefully set the scanning time in order to track more than \unit{98}\% of colloids between two successive time steps.

From direct confocal measurement~\cite{Royall2007, Poon2012}, we estimate the hard-core diameter of our colloids ($\sigma=\unit{2.75}{\micro\metre}$) and the range of the interaction potential (that confirmed our scaling of $R_g$ within $1\%$), leading to a polymer-colloid size ratio $q = 2R_g/\sigma = 0.10(6)$. Two particles are considered bonded when their centres are within $\sigma+2R_g$, which corresponds to the first minimum of $g(r)$.

>>>>

We use a novel experimental set-up that allows the observation of the entire process of gelation at a single particle level (see Methods). In a suspension of sticky colloids with long range repulsion, we suddenly turn off the repulsion. The short range attraction become dominant, eventually leading to the formation of a gel. The whole process takes place on the microscope stage and can thus be observed dynamically in three dimensions from the very beginning.

Our protocol is not equivalent to a redispersion of an existing gel at the same state point either by shaking or shear melting like in  e.g.~\cite{lu2008gelation,Teece2011,Bartlett2012}. It is closer to a quench, however we expect differences compared to a quench simulated via molecular or Brownian dynamics, because of the presence of a solvent mediating hydrodynamic interactions~\cite{Furukawa2010}. Thus our protocol might be very close to application relevant processes where the sudden addition of a component (e.g. salt, acid, depletion agent) induces aggregation in an otherwise stable suspension (e.g. milk).

Figure~\ref{fig:methods}c-h shows a computer reconstruction from experimental coordinates of a typical gelation experiment. We are able to follow the state of the system from (Fig.~\ref{fig:methods}c) a Wigner crystal where the particles are far apart due to repulsion to (Fig.~\ref{fig:methods}h) a fully percolated network. Using this protocol, we explored the phase diagram (Fig.~\ref{fig:methods}b) and found three possible behaviours: permanent gelation as above, transient gelation and cluster formation. Transient gelation is characterised by the formation of a percolating network that melts into separate clusters over a few tens of Brownian times (the time needed for an isolated particle to diffuse over its own radius, here about $\unit{10}{second}$).

We observe in Fig.~\ref{fig:methods}b that the spinodal line predicted by the generalized free volume theory~\cite{Fleer2008} corresponds roughly to the boundary between permanent and transient gelation. Furthermore, it means that in our system gelation can happen transiently without an underlying spinodal decomposition dynamics, somewhat in contradiction to~\cite{lu2008gelation}.

@@

The data were collected on a Leica SP5 confocal microscope, using \unit{532}{\nano\meter} laser excitation. The temperature was controlled on both stage and objective lens, allowing a more precise density matching. The scanned volume is at least \unit{82 \times 82 \times 85}{\micro\metre}. Standard particle tracking techniques allow us to localise our particles with errors below $1\%$ of a diameter. We carefully set the scanning time in order to be able to follow between two successive time steps all the colloids that stay in the field of view.

From direct confocal measurement~\cite{Royall2007, Poon2012}, we estimate the hard-core diameter of our colloids ($\sigma=\unit{2.75}{\micro\metre}$) and the range of the interaction potential (that confirmed our scaling of $R_g$ within $1\%$), leading to a polymer-colloid size ratio $q = 2R_g/\sigma = 0.10(6)$. Spinodal and binodal lines on Fig.~\ref{fig:methods}b are calculated from this size ratio using the generalized free volume theory~\cite{Fleer2008}. In principle two particles are considered bonded when their centres are within $\sigma+2R_g$, however resolution-dependent tracking imprecisions and systematic errors do not give a precise enough estimate of such short distance. Therefore we use the first minimum of $g(r)$ as a threshold, which is $\unit{3.40 \text{ or } 3.55}{\micro\metre}$ depending on the resolution.
